Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-03 WholePlotTrajectories.Rmd
small corrections

In tropical forests communities are shaped by a constant range of disturbance that changes the abiotic environment, as the light, heat and water fluxes, and the biotic interaction and competitive pressure.
The cornerstone of tropical forests ecology is then to understand their response to disturbance and its mechanisms [@White2001;@Chazdon2003a].
Forests response to disturbance has been largely studied through structural parameters rapid and convenient to measure, as aboveground biomass, tree height or stem density.
These structural parameters have then been sucessfully modeled and assessed the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009;@Rutishauser2016].
However the response of tree species diversity although it is determinant of ecosystems productivity, stability and functioning [@Tilman2014] and it would be most probably impacted by the biotic and abiotic changes after disturbance [@CazzollaGatti2014].

In the short-term diversity dynamics demonstrated negligible or even positive impacts of disturbance on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity at intermediate disturbance intensity [@Molino2001;@Kariuki2006a;@Berry2008a]. 
However, validations of the IDH in the long term remain scarce and mainly rely on the analysis of species richness that gives limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
More releveant monitoring would encompass communities composition, that are crucial for conservation issues, and evenness, that revealing ecological rules underlying communities' structure [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, the functional approach accounting for species biological attributes would be insightful as it reveals species fitness and ecosystems functioning [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
In that respect major functional traits related to species ecology and performance were largely adopted through a relevant framework [@Diaz2005;@Villeger2008a].
The functonial trait-based approach, for example, highlighted in tropical rainforests the environmental filters fostering disturbance resistant species with rapid growth and efficient resources acquisition [@Molino2001;@Haddad2008].
This translated by shifts from "conservative" slow-growing species dealing with scarce resources, to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
It was mirrored by shifts in key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction life history traits (seed mass and maximum height) [@Wright2004;@Westoby2006a;@Chave2009b].
A proper monitoring of communities response should therefore encompass taxonomic and functional diversity and composition measures to test the validity of the IDH in the long term for tropical forests, and clarify the resilience of communities evenness, composition,and also functioning.
The trajectories followed by all these facets would highlight the role of deterministic processes, like competitive exclusion or abiotic selection, and the communities' convergence maintaining intrinsinc differences in diversity and composition,which is as much insgights for future adaptive conservation strategies [@Adler2007].

@@

Paracou station in French Guiana (5°18'N and 52°53'W) is located in a lowland tropical rain forest corresponding to a tropical wet climate with mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) with a 3-months dry season (< 100 mm.months^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. 
Elevation ranges between 5 and 50 m and mean annual temperature is 26°C. 
Soils correspond to thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains.
The experiment corresponds to a network of twelve 6.25ha plots that have undergone a gradient of three logging, thinning and fuelwood treatments (Table \@ref(tab:Tab1)).
Disturbance treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Blanc2009].

@@

The study site corresponds to a tropical rainforest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical families. 
In the twelve experimental plots of the experiment, all trees above 10 cm DBH were mapped and measured annually since 1984. 
During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas other less common species were identified under two identifiers only separating trees and palm trees. 
The botanical campaigns carried every 5 to 6 years to identify all trees at the species level only started in 2003 and identification practices varied among plots and successive campaigns. 
This raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that had to be propagated to composition and diversity metrics throught a Bayesian framework. 
The uncertainty propagation was done by the replenishment of inventories completed at genus level from real incomplete ones on the basisof vernacular/botanical names association.

Vernacular names were replaced through multinomial trials $M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)$ based on the observed association probability $\big[\alpha_1, \alpha_2,…, \alpha_3\big]$ between each vernacular name *v* and the species $\big[s_1, s_2, …, s_N\big]$ recorded in the inventory.

To avoid remaining identification caveats and consider complete inventories, the simulated inventories were then reported at genus.
See appendix 1 and @Aubry-Kientz2013 for the detailed methodology.

@@

The trait database came from the BRIDGE project [^1] where traits values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou. 
Missing trait values were filled using multivariate imputation by chained equation (mice) restricted to samples pertaining to the next higher taxonomic level, in order to account for the phylogenetic signal of the functional traits. 
The dataset comprised 294 botanical species pertaining to 157 botanical genera.

@@

[^1] http://www.ecofog.gf/Bridge/

@@

To counter the remaining taxonomic uncertainty plots taxonomic composition and diversity were analysed at the genus level, *i.e.* referring to the genus of observed or trialed botanical names.
Variations in plots taxonomic and functional composition after disturbance was visualized by their trajectories in a two-dimensional ordination space over 30 years.
Two NMDS were conducted, either from taxonomic flora inventories or from plots functional composition based on the 8 leaf, stem and life history traits. 
Plots trajectories along time was reported comparatively to the inventories in 1989, 5 years after disturbance, which corresponded to first inventory with a sufficient degree of uncertainty (<30% of undetermined trees). 
The inventories dissimilarity compared to the reference 1989 inventory was reported using occurrence-based (Jaccard) and abundance-based (Bray-Curtis) similarity measures. 
The trajectory of inventories along time was visualized with the euclidean distance in the two-dimensional ordination space to the 1989 inventory.
The trajectories of the leaf and stem and life traits were also visualized with the community weighted means (CWM), representing the average trait value in a community weighted by relative abundance of the species carrying each value [@Diaz2007;@Garnier2004]. 
To compensate the intrinsinc difference among plots the trajectories corresponded to the differences along time with the reference inventory in 1984 (5 year after disturbance).
Species seed mass corresponded to 5 classes of increasing mass, seed mass trajectories were  therefore reported as the proportion of each class in the inventories.

@@

Over time, 828388 trees and 591 botanical species pertaining to 223 genus and 64 botanical families were recorded. 
Taxonomic and functional plots trajectories were examined through the ordination in a two dimensional ordiantion space of successive inventories from 1989 (5 years after disturbance) to 2015 (31 years after disturbance).
Classifications were performed using either abundance-based Bray-Curtis (Figure \@ref(fig:NMDSplans)) or incidence-based Jaccard dissimilarity, this last having given similar results (data not shown).

@@

The changes observed in plots functional composition went hand to hand with consistent trajectories of the 8 functional and life history traits (Leaf thickness, chlorophyll content, toughness and specific area,wood specific gravity and barck thickness and seed mass and maximum adult height) visaualized with plots community weighted means (CWM) (Figure \@ref(fig:CWM)).

Except for leaf cholorphyll content, which continued to increase for some T3 and T2 plots 30 years after disturbance, all traits and seed mass proportions displayed a unimodal trajectories.
The weighted means of communities specific maximum height at adult stage (*Hmax*), leaf toughness (*L_toughness*) and wood specific gravity (*WD*) remained significantly lower than their initial value and than these of the control plots (Figure \@ref(fig:CWM)).
The weighted means of bark thickness (*Bark_thick*) similarly remained substantilly higher than initially for all disturbed plots while the specfic leaf area (*SLA*) had almost recovered its initial value and this of the undisturbed plots at the end of the experiment.
Once again the maximum difference to initial state was correlated to the disturbance intensity ($\rho_{spearman}^{L_thickness}=0.67$, $\rho_{spearman}^{L_chloro}=0.45$, $\rho_{spearman}^{L_toughness}=-0.43$, $\rho_{spearman}^{SLA}=0.93$, $\rho_{spearman}^{WD}=-0.78$, $\rho_{spearman}^{Bark-thickness}=0.88$, $\rho_{spearman}^{Hmax}=-0.48$).

```{r CWM, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, (ref:capThickness), chlorophyll content, (ref:capChloro), toughness, (ref:capThoughness) and specific area, (ref:capSLA)), 2 stem traits (wood specific gravity, (ref:capWD), and bark thickness, (ref:capBarkThick)) and one life trait (Specific maximum height at adult stage, (ref:capHmax)). Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure*"}
load("DB/CWM")

@@

For undisturbed plots the Richness, Shannon and Simpson diversity remained stable over the 30 monitored years.
After disturbance the richness increased when the disturbance intensity was low, with a maximum increase of 14 botanical genuses (plot 3 from treatment 2).
When the disturbance was intense however, plot staxonomic richness followed a unimodal decrease with a return to initial values after intense disturbance.
The taxonomic evenness (Shannon and Simpson diversities) significantly increased whenever the disturbance intensity.
The evenness followed a unimodal trajectory with a just beginning return towards initial values and a maximum reached around 20 years and positively correlated to the disturbance intensity ($\rho_{spearman}^{Shannon}=0.86$, and $\rho_{spearman}^{Simpson}=0.89$).
Only two T3 plots, plots 8 and 12, remained increasing 30 years after disturbance \@ref(fig:DivTaxo)), suggesting a similar be much delayed unimodal trajectory.

@@

The functional diversity trajectories of plots based on the 8 leaf, stem and life history traits was examined through the Rao diversity.
For all undisturbed plots the functional diversity remained comparable to the reference values 5 years after disturbance (1989 inventories).
For all disturbed plots the functional trajectories followed a unimodal trajectory with a maximum positively correlated to disturbance intensity ($\rho_{spearman}=0.73$) (Figure \@ref(fig:DivFun)). 
The plot 7 from treatment 1 that displayed a constantly outlying diversity was removed from the graphical representation for better readability, see appendix for full graph.
Thirty years after disturbance all plots, whenever the initial disturbance intensity, regained diversity values similar to their initial avule and to those of control plots.

```{r DivFun, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes.", fig.env="figure"}
load("DB/FunctionalTraj_ForGraphs")

@@

The monitoring of disturbed forest communities confirmed a limited negative impact of disturbance on species richness, as observed on several post logging surveys [@Cannon1998;@Baraloto2012a].
Thirty years after disturbance, the genus richness was restored to initial plot values after high disturbance intensity while it substantially increased after low disturbance intensity, reaching a gain of almost 12 genuses for some plots.

@@

The trajectories of almost all leaf and stem functional traits, life history traits,and functional diversity followed a hump-back trajectory which maximum were positively correlated to the disturbance intensity, and were recovering towards initial values after 30 years that trajectory. 
Functoinal traits and communities functional diversity are the most direct link between biodiversity and ecosystem functioning [@Diaz2005]: these trajectories reflected a consistent recovery in the long term of ecosystem processes [@Guariguata2001].

@@

Despite the consistent recovery of initial diversity and composition there was a time lag between the trajectories followed by taxonomic and functional characteristics.
Taxonomic composition had longer time path and while communities had recovered their functional characteristics, their taxonomic composition and evenness remained altered by disturbance.
This delay between functional and taxonomic dynamics was already observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015;@Guariguata2001].
The functional diversity of communities rely on the functional strategy of dominant species, according to the "vegetation quantity effect" [@Grime1998].
The functional trajectories of communities are then driven first by the increase of diversity and evenness of dominant species and then by the recruitment of species resembling the old pre-disturbance community which reduced the overall diversity.
Although dominant species restoredthe functional pre-disturbance characteristics, communities composition remained altered and the infrequent species of initial communities remain missing to achieve plots recovery.
According to the funcitonal traits trajectories, those infrequent species would match the dominant functional type and would participate to the functional redundancy.
The functional redundancy, thas is the functional overlap between species that is typical of the huge biodiversity of tropical forests [@Bellwood2006], therefore was not restored 30 years after disturbance and this is majr to consider as it defines forests' resilience [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Besides the long-term alteration of functional redundancy, the persistent compositional changes which might have favored disturbance resistant species [@Haddad2008], lianas or epiphytes [@Martin2013] and would probably have come along with environmental change, like in the soils nutrient cycling and compaction [@Olander2005].
These presistent changes highly question forest's resilience[@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

>>>>

In tropical forests communities are shaped by a constant range of disturbance that changes the abiotic environment, as the light, the heat and water fluxes, and the biotic interaction and competitive pressure.
The cornerstone of tropical forests ecology is then to understand their response to disturbance and its mechanisms [@White2001;@Chazdon2003a].
Forests response has been largely studied through structural parameters rapid and convenient to measure, as aboveground biomass, tree height or stem density.
These structural parameters have then been sucessfully modeled and allowed to assess the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009;@Rutishauser2016].
However the response of tree species diversity, which is determinant of ecosystems productivity, stability and functioning [@Tilman2014] and would be most probably impacted by post-disturbance environmental changes [@CazzollaGatti2014], remains unclear.

In the short-term diversity dynamics demonstrated negligible or even positive impacts of disturbance on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity at intermediate disturbance intensity [@Molino2001;@Kariuki2006a;@Berry2008a]. 
However, validations of the IDH in the long term remain scarce and mainly rely on the analysis of species richness that gives limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
More releveant monitoring would encompass communities composition, that is crucial for conservation issues, and evenness, that reveals ecological rules underlying communities' structure [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, the functional approach accounting for species biological attributes would be insightful as it reveals species fitness and ecosystems functioning [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
In that respect major functional traits related to species ecology and performance were largely adopted through a relevant framework [@Diaz2005;@Villeger2008a].
The functional trait-based approach, for example, highlighted in tropical rainforests the environmental filters fostering disturbance resistant species with rapid growth and efficient resources acquisition [@Molino2001;@Haddad2008].
This translated by shifts from "conservative" slow-growing species dealing with scarce resources, to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
It was mirrored by shifts in key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction life history traits (seed mass and maximum height) [@Wright2004;@Westoby2006a;@Chave2009b].
A proper monitoring of communities response should therefore encompass taxonomic and functional diversity and composition measures to test the validity of the IDH in the long term for tropical forests, and clarify the resilience of communities evenness, composition, and also functioning.
The trajectories followed by all these facets would highlight the role of deterministic processes, like competitive exclusion or abiotic selection, and the communities' convergence maintaining intrinsinc differences in diversity and composition,which is as much insgights for future adaptive conservation strategies [@Adler2007].

@@

Paracou station in French Guiana (5°18'N and 52°53'W) is located in a lowland tropical rain forest corresponding to a tropical wet climate with mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) with a 3-month dry season (< 100 mm.month^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. 
Elevation ranges between 5 and 50 m and mean annual temperature is 26°C. 
Soils correspond to thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains.
The experiment corresponds to a network of twelve 6.25ha plots that have undergone a gradient of three logging, thinning and fuelwood treatments (Table \@ref(tab:Tab1)).
Disturbance treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Blanc2009].

@@

The study site corresponds to a tropical rainforest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical families. 
In the twelve experimental plots of the experiment, all trees above 10 cm DBH were mapped and measured annually since 1984. 
During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas other less common species were identified under two identifiers only separating trees and palm trees. 
The botanical campaigns carried every 5 to 6 years to identify all trees at the species level only started in 2003 and identification practices varied among plots and successive campaigns. 
This raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that had to be propagated to composition and diversity metrics through a Bayesian framework. 
The uncertainty propagation was done by the replenishment of inventories completed at genus level from real incomplete ones on the basis of vernacular/botanical names association.

Vernacular names were replaced through multinomial trials $M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)$ based on the observed association probability $\big[\alpha_1, \alpha_2,…, \alpha_3\big]$ between each vernacular name *v* and the species $\big[s_1, s_2, …, s_N\big]$ recorded in the inventory. 
See appendix 1 and @Aubry-Kientz2013 for the detailed methodology.

@@

The trait database came from the BRIDGE project [^1] where trait values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou. 
Missing trait values were filled using multivariate imputation by chained equation (mice) restricted to samples pertaining to the next higher taxonomic level, in order to account for the phylogenetic signal of the functional traits. 
The dataset comprised 294 botanical species pertaining to 157 botanical genera.

@@

[^1]: http://www.ecofog.gf/Bridge/

@@

To counter the remaining taxonomic uncertainty, plots taxonomic composition and diversity were analysed at the genus level, *i.e.* referring to the genus of observed or trialed botanical names.
Variations in plots taxonomic and functional composition after disturbance was visualized by their trajectories in a two-dimensional ordination space over 30 years.
Two NMDS were conducted, either from taxonomic flora inventories or from plots functional composition based on the 8 leaf, stem and life history traits. 
Plots trajectories along time was reported comparatively to the inventories in 1989, 5 years after disturbance, which corresponded to first inventory with a sufficient degree of uncertainty (<30% of undetermined trees). 
The inventories dissimilarity compared to the reference 1989 inventory was reported using occurrence-based (Jaccard) and abundance-based (Bray-Curtis) similarity measures. 
The trajectory of inventories along time was visualized with the euclidean distance in the two-dimensional ordination space to the 1989 inventory.
The trajectories of the leaf and stem and life traits were also visualized with the community weighted means (CWM), representing the average trait value in a community weighted by relative abundance of the species carrying each value [@Diaz2007;@Garnier2004]. 
To compensate the intrinsinc difference among plots the trajectories corresponded to the differences along time with the reference inventory in 1989.
Species seed mass corresponded to 5 classes of increasing mass, seed mass trajectories were  therefore reported as the proportion of each class in the inventories.

@@

Over time, 828388 trees and 591 botanical species pertaining to 223 genus and 64 botanical families were recorded. 
Taxonomic and functional plots trajectories were examined through the trajecotries in a two dimensional ordination space of successive inventories from 1989 (5 years after disturbance) to 2015 (31 years after disturbance).
Classifications were performed using either abundance-based Bray-Curtis (Figure \@ref(fig:NMDSplans)) or incidence-based Jaccard dissimilarity, this last having given similar results (data not shown).

@@

The changes observed in plots functional composition went hand to hand with consistent trajectories of the 8 functional and life history traits (Leaf thickness, chlorophyll content, toughness and specific area, wood specific gravity and bark thickness and seed mass and maximum adult height) visualized with plots community weighted means (CWM) (Figure \@ref(fig:CWM)).

Except for leaf cholorphyll content, which continued to increase for some T3 and T2 plots 30 years after disturbance, all traits and seed mass proportions displayed a unimodal trajectories.
The weighted means of communities specific maximum height at adult stage (*Hmax*), leaf toughness (*L_toughness*) and wood specific gravity (*WD*) remained significantly lower than their initial value and than these of the control plots (Figure \@ref(fig:CWM)).
The weighted means of bark thickness (*Bark_thick*) similarly remained substantilly higher than initially for all disturbed plots while the specfic leaf area (*SLA*) had almost recovered its initial value and this of the undisturbed plots at the end of the experiment.
Once again the maximum difference to initial state was correlated to the disturbance intensity ($\rho_{spearman}^{L_{thickness}}=0.67$, $\rho_{spearman}^{L_{chloro}}=0.45$, $\rho_{spearman}^{L_{toughness}}=-0.43$, $\rho_{spearman}^{SLA}=0.93$, $\rho_{spearman}^{WD}=-0.78$, $\rho_{spearman}^{Bark-thickness}=0.88$, $\rho_{spearman}^{Hmax}=-0.48$).

```{r CWM, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, (ref:capThickness), chlorophyll content, (ref:capChloro), toughness, (ref:capThoughness) and specific area, (ref:capSLA)), 2 stem traits (wood specific gravity, (ref:capWD), and bark thickness, (ref:capBarkThick)) and one life trait (Specific maximum height at adult stage, (ref:capHmax)). Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure*"}
load("DB/CWM")

@@

For undisturbed plots the Richness, Shannon and Simpson diversities remained stable over the 30 monitored years.
After disturbance the richness increased when the disturbance intensity was low, with a maximum increase of 14 botanical genera (plot 3 from treatment 2).
When the disturbance was intense however, plots taxonomic richness followed a unimodal decrease with a return to initial values after intense disturbance.
The taxonomic evenness (Shannon and Simpson diversities) significantly increased irrespective of the disturbance intensity.
The evenness followed a unimodal trajectory with a just beginning return towards initial values and a maximum reached around 20 years and positively correlated to the disturbance intensity ($\rho_{spearman}^{Shannon}=0.86$, and $\rho_{spearman}^{Simpson}=0.89$).
Only two T3 plots, plots 8 and 12, remained increasing 30 years after disturbance \@ref(fig:DivTaxo)), suggesting a similar be much delayed unimodal trajectory.

@@

The functional diversity trajectories of plots based on the 8 leaf, stem and life history traits was examined through the Rao diversity.
For all undisturbed plots the functional diversity remained comparable to the reference values 5 years after disturbance (1989 inventories).
For all disturbed plots the functional trajectories followed a unimodal trajectory with a maximum positively correlated to disturbance intensity ($\rho_{spearman}=0.73$) (Figure \@ref(fig:DivFun)). 
The plot 7 from treatment 1 that displayed a constantly outlying diversity was removed from the graphical representation for better readability, see appendix for full graph.
Thirty years after disturbance all plots, whenever the initial disturbance intensity, regained diversity values similar to their initial value and to those of control plots.

```{r DivFun, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure"}
load("DB/FunctionalTraj_ForGraphs")

@@

The monitoring of disturbed forest communities confirmed a limited negative impact of disturbance on species richness, as observed on several post logging surveys [@Cannon1998;@Baraloto2012a].
Thirty years after disturbance, the genus richness was restored to initial plot values after high disturbance intensity while it substantially increased after low disturbance intensity, reaching a gain of almost 12 genera for some plots.

@@

The trajectories of almost all leaf and stem functional traits, life history traits, and functional diversity followed hump-back trajectories whose maximum were positively correlated to the disturbance intensity, and were recovering towards initial values after 30 years. 
Functional traits and communities functional diversity are the most direct link between biodiversity and ecosystem functioning [@Diaz2005]: these trajectories reflected a consistent recovery in the long term of ecosystem processes [@Guariguata2001].

@@

Despite the consistent recovery of initial diversity and composition there was a time lag between the trajectories followed by taxonomic and functional characteristics.
Taxonomic composition had longer time path and while communities had recovered their functional characteristics, their taxonomic composition and evenness remained altered by disturbance.

This delay between functional and taxonomic dynamics was already observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015;@Guariguata2001].
The functional diversity of communities rely on the functional strategy of dominant species, according to the "vegetation quantity effect" [@Grime1998].
The functional trajectories of communities are then driven first by the increase of diversity and evenness of dominant species and then by the recruitment of species resembling the old pre-disturbance community which reduced the overall diversity.
Although dominant species restored the functional pre-disturbance characteristics, communities composition remained altered and the infrequent species of initial communities remain missing to achieve plots recovery.
According to the functional traits trajectories, those infrequent species would match the dominant functional type and would participate to the functional redundancy.
The functional redundancy, thas is the functional overlap between species that is typical of the huge biodiversity of tropical forests [@Bellwood2006], therefore was not restored 30 years after disturbance and this is major to consider as it defines forests' resilience [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Besides the long-term alteration of functional redundancy, there was probably persistent compositional changes favoring disturbance resistant species [@Haddad2008], lianas or epiphytes [@Martin2013] and environmental changes, like in the soils nutrient cycling and compaction [@Olander2005].
These presistent changes highly question forest's resilience [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].
